preciseTAD: A machine learning framework for precise TAD boundary prediction at base pair resolution

### 2020-07-06 Manuscript/manuscript.Rmd
Edits through Figure 5

## 5 kb resolution, random under-sampling, and distance-based predictors provided optimal performance for TAD boundary prediction

Expectedly, when using data with class imbalance present, that is, no re-sampling, the models exhibited low balanced accuracies, with very little variability among different resolutions (Figure 3). Similarly, poor performances were found when using ROS as well. However, RUS and SMOTE re-sampling led to drastic improvement in performance, especially at higher resolutions. We found that RUS marginally outperformed SMOTE as the most optimal class balancing technique for all resolutions and predictor types. These results suggest that simple random under-sampling alone can drastically improve TAD boundary region prediction.

Additionally, we found that at higher resolutions, random forest classifiers built using a distance-type predictor space yielded substantially higher balanced accuracies when compared to the peak signal stength, overlap count, and overlap percent predictor types. As with comparing class balancing techniques, this improvement was less evident at lower resolutions. These results were found to be consistent for the K562 cell line (Supplementary Figure 2). Random forest models built on PEAKACHU loop boundaries with SMOTE resampling only marginally outperformed RUS for both GM12878 and K562 cell lines (Supplementary Figure 3). Five kilobase resolution genomic bins was found to result in optimal prediction for both TAD and loop boundaries, on both cell lines. Our results indicate that random under-sampling, distance-type predictors, and Hi-C data at 5kb resolution are the optimal set of data level characteristics for both TAD and chromatin loop boundary region prediction.

@@

![**???[needs updating with new framework, takes longest, do last]Models built using distance-type predictors combined with random under-sampling at 5 kb yielded optimal TAD boundary region predictive performance for GM12878.** Line plots comparing the performances of random forest classifiers built using different combinations of data level characteristics. Performances are compared across resolution, within each predictor-type: (A) Signal, (B) OC, (C) OP, and (D) Distance, and between each re-sampling technique: no re-sampling (None; red), random over-sampling (ROS; green), random under-sampling (RUS; blue), and synthetic minority over-sampling (SMOTE; purple). Performances were evaluated using balanced accuracy.](Figures/figure3_performances.png)

## Transcription factor binding sites outperform HM- and BroadHMM-specific models

We hypothesized that the type of genomic annotations may affect the predictive performance. Using the aforementioned optimal settings, we built separate random forest models using BroadHMM chromatin segmentations, histone modifications (HM), and transcription factor binding sites (TFBS). We found that models built on TFBSs outperformed other annotation-specific models, and these results were consistent for loop boundaries, for both cell lines (Figure 4A; Supplementary Figures 4A). These results suggest that TFBS are the primary drivers of TAD and loop boundary formation in both the GM12878 and K562 cell lines.

@@

![**Assessing top ranked transcription factors in TAD and loop boundary region prediction for GM12878.** (A) Barplots comparing performances of TAD (ARROWHEAD) and loop (PEAKACHU) boundary region prediction models using distance-type specific classes of functional genomic elements as the predictor space including histone modifications (HM), cis-regulated elements (BroadHMM), transcription factor binding sites (TFBS), in addition to a model containing all three classes (ALL). (B) Recursive feature elimination (RFE) analysis used to reduce the number of TFBS to include in the optimally built random forest model. Error bars represent standard deviation from the cross-validated balance accuracy across each autosomal chromosome. (C) Clustered heatmap of predictive importances for the union of the top 8 most predictive TFBS across chromosomes as determined from performing RFE.](Figures/figure4_annots_specifc_rfe_importances_arrowhead_peakachu.png)

@@

After obtaining our optimally built random forest model, we investigated whether we could leverage our model to alleviate resolution limitations of Hi-C data by predicting TAD boundaries at base pair resolution. Our method, referred to as *preciseTAD*, leverages a user specified binary classifier with a combination of density-based hierarchical clustering and a scalable version of partitioning around mediods (PAM), known as clustering large applications (CLARA) to precisely identify the most representative and biologically meaningful predicted chromosome-specific domain boundaries (see Methods).

When built using ARROWHEAD ground truth boundaries at 5 kb, *preciseTAD* was able to predict a total of ??? and ??? boundaries for GM12878 and K562 cell lines respectively. We reported less predicted TAD boundaries than ARROWHEAD (Table 1, Supplementary Table 3). This is attributed to ARROWHEAD's inherent inflation of called TAD boundaries, especially near fuzzy areas along the diagonal of Hi-C contact matrices (Figure 5A). However, under the framework of *preciseTAD*, any two TADs are inherently conjoined by only one predicted TAD boundary, leading to comparatively less TAD boundaries. Visually speaking, we also found that predicted TAD boundaries identified by *preciseTAD* were much more likely to demarcate high concentrations of inter-chromosomal contacts as seen on a Hi-C contact matrix (Figure 5A). [???add peakachu totals when ready]

Next, we evaluated the biological significance of *preciseTAD* boundaries. Enriched heatmaps indicated that our predicted boundaries were more enriched for CTCF sites than either ARROWHEAD or PEAKACHU boundaries alone (Figure 5 B-C). Likewise, *preciseTAD* boundaries were also more enriched for other molecular drivers of 3D chromatin including RAD21, SMC3, and ZNF143 for both GM12878 and K562 cell lines (Supplementary Figures 8-9). When compared to called TAD or loop boundaries, *preciseTAD* boundaries were found to be statistically significantly closer to the same cell line-specific top ranked TFBS (p=??? and p=??? vs ARROWHEAD and PEAKACHU respectively) (Figure 5D, Supplementary Figure 10-11). We also inspected the affect of resolution on locations of our predicted TAD boundaries. We found that *preciseTAD* boundaries exhibited higher pairwise Jaccard indices between adjacent resolutions than ARROWHEAD boundaries, thus implying stronger invariance toward resolution (Figure 5E).

![**preciseTAD boundaries are enriched in CTCF sites.** (A) Visualizing locations of predicted vs called ARROWHEAD TAD boundaries on GM12878 (chr 1: 70600000-71530000). The upper triangular contact map shows the normalized number of contacts within the specified window using the JUICER toolbox. The blue vertical lines represent boundaries identified by ARROWHEAD, with the corresponding called TADs outlined in blue on the upper diagonal contact matrix above. The black line represents the predicted probability of each base pair being a TAD boundary when validated by our random forest. The green vertical lines represent the *preciseTAD boundary points* (PTBP) predicted by *preciseTAD*, with the corresponding demarcated TADs outlined above in green. Below is a zoomed in portion of the genome showing a cluster of base pairs with high probability representing a single *preciseTAD boundary region (PTBR) colored in yellow. (B) Enriched heatmaps comparing the enrichment of CTCF sites around 5 kb flanked preciseTAD and ARROWHEAD TAD boundaries respectively. (C) Enriched heatmaps comparing the enrichment of CTCF sites around 10 kb flanked preciseTAD and PEAKACHU loop boundaries respectively. (D) $Log_{2}$ genomic distances from boundaries to nearest CTCF site between preciseTAD boundaries and both ARROWHEAD and PEAKACHU boundaries, respectively. (E) A atrix of pairwise Jaccard indices across resolutions shows more invariance toward resolution for *preciseTAD* (upper triangular) compared to ARROWHEAD (lower triangular).???update when ready](Figures/figure5_enrichment_and_log2dist_ctcf.png)

@@

**Supplementary Figure 1.** Density curves illustrate the normalization of distances before (red) and after (blue) performing a $log_{2}$ transformation across differing bin sizes including 10 kb, 25 kb, 50 kb, and 100 kb for both the (A) GM12878 and (B) K562 cell lines respectively. Distances are measured as the number of base pairs from the center of a genomic bin to the center of the nearest genomic annotation. Each density curve represents a particular genomic annotation.

@@

**Supplementary Figure 2. Models built using distance-type predictors combined with random under-sampling at 5 kb yielded optimal predictive performance for K562.** Line plots comparing the performances of random forest classifiers built using ARROWHEAD boundaries at different combinations of data level characteristics. Performances are compared across resolution, within each predictor-type: (A) Signal (B) OC, (C) OP, and (D) Distance, and between each re-sampling technique: no re-sampling (None; red), random over-sampling (ROS; green), random under-sampling (RUS; blue), and synthetic minority over-sampling (SMOTE; purple). Performances were evaluated using balanced accuracy.

**Supplementary Figure 3. Models built using distance-type predictors combined with SMOTE re-sampling yielded optimal predictive performance for GM12878 K562.** Line plots comparing the performances of random forest classifiers built using using PEAKACHU boundaries at different combinations of data level characteristics for (A) GM12878 and (B) K562. Performances are compared across resampling technique: no re-sampling (None; red), random over-sampling (ROS; green), random under-sampling (RUS; blue), and synthetic minority over-sampling (SMOTE; purple), and within each predictor-type: Signal, OC, OP, and Distance.

**Supplementary Figure 4. Assessing top ranked transcription factors in TAD and loop boundary prediction for K562.** (A) Barplots comparing performances of TAD (ARROWHEAD) and loop (PEAKACHU) boundary prediction models using specific classes of functional genomic elements as the predictor space including histone modifications (HM), cis-regulated elements (BroadHMM), transcription factor binding sites (TFBS), in addition to a model containing all three classes (ALL). (B) Recursive feature elimination (RFE) analysis used to reduce the number of TFBS to include in the optimally built random forest model. Error bars represent standard deviation from the cross-validated accuracy across each chromosome. (C) Clustered heatmap of predictive importances for the union of the top 8 most predictive TFBS across chromosomes as determined from performing RFE.

>>>>

## Random under-sampling, distance-based predictors, and high-resolution Hi-C data provide optimal performance for boundary prediction

Expectedly, when using data with class imbalance present, that is, no re-sampling, the models exhibited low balanced accuracies, with very little variability among different resolutions (Figure 3). Similarly, poor performances were found when using ROS as well. However, RUS and SMOTE re-sampling led to drastic improvement in performance, especially at higher resolutions. We found that RUS marginally outperformed SMOTE as the most optimal class balancing technique for all resolutions and predictor types. These results suggest that simple random under-sampling can significantly improve TAD boundary region prediction.

Additionally, we found that using a distance-type predictor space yielded substantially higher balanced accuracies when compared to the peak signal stength, overlap count, and overlap percent predictor types. As with class balancing techniques, this improvement was less evident at lower resolutions. These results were consistent for the K562 cell line (Supplementary Figure 2). Random forest models built on PEAKACHU loop boundaries with SMOTE resampling only marginally outperformed RUS for both GM12878 and K562 cell lines (Supplementary Figure 3). Five kilobase resolution genomic bins was found to result in optimal prediction for both TAD and loop boundaries, on both cell lines. Our results indicate that random under-sampling, distance-type predictors, and high-resolution Hi-C data provide the optimal set of data level characteristics for both TAD and loop boundary prediction.

@@

![**???[needs updating with new framework, takes longest, do last] Distance-type predictors combined with random under-sampling yielded optimal TAD boundary region predictive performance for GM12879 ARROWHEAD TAD boundaries.** Balanced accuracy is compared across resolution, within each predictor-type: (A) Signal, (B) OC, (C) OP, and (D) Distance, and across re-sampling techniques: no re-sampling (None; red), random over-sampling (ROS; green), random under-sampling (RUS; blue), and synthetic minority over-sampling (SMOTE; purple).](Figures/figure3_performances.png)

## Transcription factor binding sites outperform histone- and chromatin segmentation-specific models

We hypothesized that the type of genomic annotations may also affect the predictive performance. Using the established optimal settings (RUS, Distance), we built separate random forest models using BroadHMM chromatin segmentations, histone modifications (HM), and transcription factor binding sites (TFBS). We found that models built on TFBSs outperformed other annotation-specific models, and these results were consistent for loop boundaries, for both cell lines (Figure 4A; Supplementary Figures 4A). These results suggest that TFBS are the primary drivers of TAD and loop boundary formation in both the GM12878 and K562 cell lines.

@@

![**Binding of SMC3, Rad21, CTCF, and Znf143 transcription factors accurately predicts TAD and loop boundaries in GM12878 cell line.** (A) Barplots comparing performances of TAD (ARROWHEAD) and loop (PEAKACHU) boundary prediction models using histone modifications (HM), cis-regulated elements (BroadHMM), transcription factor binding sites (TFBS), in addition to a model containing all three classes (ALL). (B) Recursive feature elimination (RFE) analysis used to select the optimal number of predictors. Error bars represent standard deviation from the cross-validated accuracy across each chromosome. (C) Clustered heatmap of the predictive importance for the union of the top 8 chromosome-specific most predictive TFBSs.](Figures/figure4_annots_specifc_rfe_importances_arrowhead_peakachu.png)

@@

Using our optimally built random forest model, we attempted to overcome resolution limitations of Hi-C data by predicting TAD [??? Check throughout the manuscript - we may need to define "interacting domains" referring to TADs and loops] boundaries at base resolution. Intuitively, the predictor-response space is built on a base-resolution level, and the model trained on a bin-level space predicts the probability of each base being a boundary. Our method, referred to as *preciseTAD*, leverages a user-specified [??? We need pre-trained models and understanding how they perform on different cell lines] model with density-based hierarchical clustering (DBSCAN) and a scalable partitioning around mediods clustering (CLARA) to predict the boundary location ad base resolution (see Methods).

When built using ARROWHEAD ground truth boundaries at 5 kb, *preciseTAD* was able to predict a total of ??? and ??? boundaries for GM12878 and K562 cell lines respectively. We reported less predicted TAD boundaries than ARROWHEAD (Table 1, Supplementary Table 3) [??? Where is this info?]. This may be attributed to ARROWHEAD's boundaries frequently being separated by gaps, the effect especially pronounced near sparse areas of Hi-C contact matrices (Figure 5A) [??? I don't see it, data not shown?]. However, under the framework of *preciseTAD*, any two TADs are inherently conjoined by only one predicted TAD boundary, leading to comparatively less TAD boundaries. [??? Unclear explanation, so what? preciseTAD is driven by genomic annotations. May it be too stringent thresholds? This all may be for the discussion, if we have a good understanding why] We also found that predicted TAD boundaries identified by *preciseTAD* were visually more likely to demarcate high concentrations of inter-chromosomal contacts as seen on a Hi-C contact matrix (Figure 5A). [???add peakachu totals when ready]

Next, we evaluated the biological significance of *preciseTAD* boundaries. Signal heatmaps [??? Make sure to mention DeepTools in Methods] indicated that our predicted boundaries were more enriched for CTCF sites than either ARROWHEAD or PEAKACHU boundaries alone (Figure 5 B-C). Likewise, *preciseTAD* boundaries were also more enriched for other molecular drivers of 3D chromatin including RAD21, SMC3, and ZNF143 for both GM12878 and K562 cell lines (Supplementary Figures 8-9). When compared to called TAD or loop boundaries, *preciseTAD* boundaries were found to be statistically significantly closer to the same cell line-specific top ranked TFBS (p=??? and p=??? vs ARROWHEAD and PEAKACHU respectively) (Figure 5D, Supplementary Figure 10-11). We also inspected the effect of resolution on locations of our predicted TAD boundaries. We found that *preciseTAD* boundaries exhibited higher pairwise Jaccard indices between adjacent resolutions than ARROWHEAD boundaries, thus implying stronger invariance toward resolution (Figure 5E).

![**preciseTAD boundaries are enriched in CTCF sites.** (A) The location of preciseTAD (green) vs. ARROWHEAD (blue) boundaries on GM12878 (chr 1: 70600000-71530000). The black line represents the probability of each base being a TAD boundary. The green vertical lines represent the *preciseTAD boundary points* (PTBP). A zoomed-in portion of the genome shows a yellow-highlighted cluster of bases with high probability of being a boundary, the *preciseTAD boundary region (PTBR). (B) Heatmaps comparing the enrichment of CTCF signal around preciseTAD and ARROWHEAD TAD boundaries, and (C) around preciseTAD and PEAKACHU loop boundaries. (D) $Log_{2}$ genomic distances from boundaries to nearest CTCF sites. (E) Pairwise Jaccard indices shows better overlap among *preciseTAD*-detected boundaries across resolutions (upper triangular) as compared to ARROWHEAD (lower triangular).](Figures/figure5_enrichment_and_log2dist_ctcf.png) [??? Panel E - do we have it for Peakachu? - stretched, can be put in figure. Can you color triangulars blue/red for Arrowhead/Peakachu, green for preciseTAD?]

@@

**Supplementary Figure 1.** Density curves illustrate the normalization of distances before (red) and after (blue) performing a $log_{2}$ transformation across differing bin sizes including 10 kb, 25 kb, 50 kb, and 100 kb for both the (A) GM12878 and (B) K562 cell lines respectively. Distances are measured as the number of base pairs from the center of a genomic bin to the center of the nearest genomic annotation. Each density curve represents individual genomic annotation.

@@

**Supplementary Figure 2. Distance-type predictors combined with random under-sampling yielded optimal TAD boundary region predictive performance for K562 ARROWHEAD TAD boundaries.** Balanced accuracy is compared across resolution, within each predictor-type: (A) Signal, (B) OC, (C) OP, and (D) Distance, and across re-sampling techniques: no re-sampling (None; red), random over-sampling (ROS; green), random under-sampling (RUS; blue), and synthetic minority over-sampling (SMOTE; purple).

**Supplementary Figure 3. Distance-type predictors combined with random under-sampling or SMOTE re-sampling yielded optimal predictive performance for A) GM12878 and B) K562 Peakachu loop boundaries.** Balanced accuracy is compared within each predictor-type: Signal, OC, OP, and Distance, and across across resampling technique: no re-sampling (None; red), random over-sampling (ROS; green), random under-sampling (RUS; blue), and synthetic minority over-sampling (SMOTE; purple).

**Supplementary Figure 4. Binding of SMC3, Rad21, CTCF, and Znf143 transcription factors accurately predicts TAD and loop boundaries in K562 cell line.** (A) Barplots comparing performances of TAD (ARROWHEAD) and loop (PEAKACHU) boundary prediction models using histone modifications (HM), cis-regulated elements (BroadHMM), transcription factor binding sites (TFBS), in addition to a model containing all three classes (ALL). (B) Recursive feature elimination (RFE) analysis used to select the optimal number of predictors. Error bars represent standard deviation from the cross-validated accuracy across each chromosome. (C) Clustered heatmap of the predictive importance for the union of the top 8 chromosome-specific most predictive TFBSs.
